---
output:
  html_document:
    theme: journal
  pdf_document: default
---
Exploring Red Wines by OZGUN BALABAN
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

library(ggplot2)
library(GGally)
library(grid)
library(gridExtra)
```

```{r echo=FALSE, Load_the_Data}
setwd('/Users/ozgunbalaban')
# Load the Data
red<-read.csv('wineQualityReds.csv')
```
# Overview
In this project a sample set of red wines are analyzed for 12 different 
variables. I selected red wines as a study since I am curious in learning
what makes of a good wine. We will first start with summarizing the data
and identifying every parameter. In a data study it is important to make a
background study to understand the phenomena.


# Summary of the data
```{r summary_of_the_Data}
summary(red)
```
 Analyzing the data, we have 1599 observations with 13 variables. These are;  
**X** -> unique id  
**fixed.acidity** -> acid produced in the body from sources other than CO2  
**volatile.acidity** -> acids produced by yeast and bacteria  
**citric.acid** -> rarely found in wine, can be used to boost wine acidity  
**residual.sugar** -> determines sweetnees of the wine  
**chlorides** -> saltiness of the wine (not a good thing)  
**free.sulfur.dioxide** -> sulfur oxide prevents bacteria growth  
**total.sulfur.dioxide** -> total amount  
**density** -> density of the wine - 1 is same as water  
**pH** -> acidity level 
          (it is alkaline if over 7 but all the values in set under 7)  
**sulphates** -> an additive to prevent wine fault  
**alcohol** -> alcohol level of the wine  
**quality** -> scores by sommeliers higher is better  

 This listing of the parameters reveals some interesting relations.
Acidity is one of the important things in wine. There are different opinions as
to what level of volatile acidity is appropriate for higher quality wine. 
Although too high a concentration is sure to leave an undesirable, 'vinegar' 
tasting wine, some wine's acetic acid levels are developed to create a more
'complex', desirable taste. The renowned 1947 Cheval Blanc is widely recognized
to contain high levels of volatile acidity.   

 Sugar level is also important in the 'taste' of the wine. Too much sugar might
 hide the complexity of the flavors. So looking on these parameters it is 
 important to find out how these affect the overall taste. Some interesting 
 questions that can be asked;  



# Univariate Plots Section

## Wine Quality

Wine quality is supplied by tasters. It can be in the range from 0 to 10.
Plotting the wine quality;
  
```{r echo=FALSE, Univariate_Plots}
ggplot(data=red, aes(x=quality)) +
  geom_histogram(binwidth = 1, color='black', fill='red') + 
  scale_x_continuous(breaks = seq(0, 10, 1),limits=c(0, 10)) +
  xlab("Wine Quality") +
  ylab("Number of Samples")

```

There is no outliers here. Minimum value is 3, maximum value is 8. We can see
most of the wines are stacked around 5-6 and a little bit of 7.


## Fixed Acidity
```{r echo=FALSE}
ggplot(data=red, aes(x=fixed.acidity)) +
  geom_histogram(binwidth = 0.25, color='black', fill='red') + 
  coord_cartesian() +
  xlab("Fixed Acidity (g / dm^3)") +
  ylab("Number of Samples") 


boxplot(red$fixed.acidity, horizontal = TRUE, main= "Fixed Acidity (g / dm^3)")

```

Here we see the mean around 8. There are few outliers in the right side of the
data. From this graph we can conclude wines are generally more acidic than 
basic.


## Volatile Acidity

```{r echo=FALSE}
ggplot(data=red, aes(x=volatile.acidity)) +
  geom_histogram(binwidth = 0.05, color='black', fill='red') + 
  coord_cartesian() +
  xlab("Volatile Acidity (g / dm^3)") +
  ylab("Number of Samples")

boxplot(red$volatile.acidity, horizontal = TRUE, 
        main= "Volatile Acidity (g / dm^3)")
```

The mean here is around 0.5. There are some outliers again in the right side.

## Citric Acid

```{r echo=FALSE}
ggplot(data=red, aes(x=citric.acid)) +
  geom_histogram(binwidth = 0.01, color='black', fill='red') + 
  coord_cartesian() +
  xlab("Citric Acid (g / dm^3)") +
  ylab("Number of Samples")

boxplot(red$citric.acid, horizontal = TRUE, main= "Citric Acid (g / dm^3)")

length(subset(red, citric.acid == 0)$citric.acid)

m <- subset(red, citric.acid != 0)

boxplot(m$citric.acid, horizontal = TRUE)

```

From the previous observation we know that citric acid is rarely used to 
 adjust acidity of wines. We have 132 wines with no citric acid. Searching on
 internet reveals that in EU using citric acid is banned. So it would be nice
 if we had the data of the origin of the wines used in this study.


## Residual Sugar


```{r}
grid.arrange(ggplot(red, aes( x = 1, y = residual.sugar ) ) + 
               geom_jitter(alpha = 0.1 )  + xlab('Residual Sugar (g / dm^3)')+
  ylab("Number of Samples") +
               geom_boxplot(alpha = 0.2, color = 'red' ) ,
             ggplot(red, aes( x = residual.sugar  ) ) + 
    xlab('Residual Sugar (g / dm^3)')+
  ylab("Number of Samples") +
                   geom_histogram(bins=30 ),ncol=2) 

```

Residual sugar has many outliers. Removing the outliers I plotted again.
 Sugar level will probably influence the taste of wine. So many outliers
 also mean that sugar level is open to varied taste and experimentation. People
 generaly have varied opinions about sweet wines.


## Chlorides

```{r echo=FALSE}
ggplot(data=red, aes(x=chlorides)) +
  geom_histogram(binwidth = 0.01, color='black', fill='red') + 
  coord_cartesian() +
  xlab("Chlorides (g / dm^3)") +
  ylab("Number of Samples")

boxplot(red$chlorides, horizontal = TRUE, main= "Chlorides (g / dm^3)")

ggplot(data = red, aes(x = chlorides)) +
  geom_histogram(binwidth = 0.001, color='black', fill='red') +
  scale_x_continuous(lim = c(quantile(red$chlorides, 0.05), 
                             quantile(red$chlorides, 0.95))) +
  xlab('Chlorides (g / dm^3)')+
  ylab("Number of Samples")

```

Chlorides have many outliers as well. Removing the extreme values, I plot it
again. Chlorides give unwanted salty smell but in turn it protects the
wine, so too much Chloride might make the wine taste bad but it might stop
spoilage.


## Free Sulfur Dioxide
```{r echo=FALSE}
ggplot(data=red, aes(x=free.sulfur.dioxide)) +
  geom_histogram(binwidth = 1, color='black', fill='red') + 
  coord_cartesian() +
  xlab("Free Sulfur Dioxide (mg / dm^3)") +
  ylab("Number of Samples")

boxplot(red$free.sulfur.dioxide, horizontal = TRUE, 
        main= "Free Sulfur Dioxide (mg / dm^3)")
```


Average Free Sulfur Dioxide is around 15 with few outliers. Though outliers
can be quite away from the mean value.

## Total Sulfur Dioxide

```{r echo=FALSE}
ggplot(data=red, aes(x=total.sulfur.dioxide)) +
  geom_histogram(binwidth = 3, color='black', fill='red') + 
  coord_cartesian() +
  xlab("Total Sulfur Dioxide (mg / dm^3)") +
  ylab("Number of Samples")

boxplot(red$total.sulfur.dioxide, horizontal = TRUE, 
        main= "Total Sulfur Dioxide (mg / dm^3)")
```


```{r}

```

Total Sulfur Dioxide has many outliers on the right hand side of the data.
And there are two points that are extremely away from the mean value.

## Density

```{r echo=FALSE}
ggplot(data=red, aes(x=density)) +
  geom_histogram(binwidth = 0.0002, color='black', fill='red') + 
  coord_cartesian() +
  xlab("Density (g / dm^3)") +
  ylab("Number of Samples")

boxplot(red$density, horizontal = TRUE, main= "Density (g / dm^3)")

```

Density values look very uniformly distributed, even the distances around the 
points are close, so I dont think it has a big effect on the taste.

## pH
```{r echo=FALSE}
ggplot(data=red, aes(x=pH)) +
  geom_histogram(binwidth = 0.02, color='black', fill='red') + 
  coord_cartesian() +
  xlab("pH") +
  ylab("Number of Samples")


boxplot(red$pH, horizontal = TRUE, main= "PH")
```



Observing Ph values we can see its mean is around 3.3 and it has few outliers.
 We can also observe since any pH lower than 7 is acidic, all the wines in this
 study is acidic. Since it is well spread towards both right and left, I wonder
 if this effects the taste.
 
 
## Sulphates

```{r echo=FALSE}
ggplot(data=red, aes(x=sulphates)) +
  geom_histogram(binwidth = 0.03, color='black', fill='red') + 
  coord_cartesian() +
  xlab("Sulphates (g / dm^3)") +
  ylab("Number of Samples")

boxplot(red$sulphates, horizontal = TRUE, main= "Sulphates (g / dm^3)")


```



Many outliers on the right side, Sulphates are naturally occuring molecules,
because of the fermentation. Too much sulpahtes might give a foul taste as
well.

## Alcohol
```{r echo=FALSE}
ggplot(data=red, aes(x=alcohol)) +
  geom_histogram(binwidth = 0.1, color='black', fill='red') + 
  coord_cartesian() +
  xlab("Alcohol (% volume)") +
  ylab("Number of Samples")

min(red$alcohol)
max(red$alcohol)
mean(red$alcohol)

boxplot(red$alcohol, horizontal = TRUE, main= "Alcohol (% volume)")

```

Observing alcohol level shows minimum alcohol level is 8.4 and maximum level 
is 14.9 with mean around 10.4. 

# Univariate Analysis


For a successful data analysis, univariate analysis of different parameters is
crucial. In this stage we can already notice there is a lot variance in 
certian variables whereas some are same for most of the wines such as 
densities. Having a univariate analysis at the beginning of a study, makes
it easier to understand the context.


### What is the structure of your dataset?
```{r}
str(red)
```
 
### What is/are the main feature(s) of interest in your dataset?

In this study I am more interested in how different factors effect taste of
wine. 


### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

Many features of the data will be related. Analizing the features with respect
to quality we find this relationship;

```{r echo=FALSE, Bivariate_Plots}
cor(x=red[,2:12], y=red$quality)
```
So Fixed Acidity, Citric Acid, Residual Sugar, Sulphates and Alcohol is
 positively corelated. Alcohol has a strong corelation. Whereas Volatile 
 Acidity,  Chlorides, Free Sulfur Dioxide, Total Sulfur Dioxide, Density and 
 pH is negatively corelated. Volatile Acidity has strong negative corelation.

### Did you create any new variables from existing variables in the dataset?

No, no new variable is created.

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?

As explained in previous section, Residual sugar and chlorides have outliers, 
those are plotted again after removing the top and bottom extremes. Citric
acid has many zero values. 

# Bivariate Plots Section


```{r echo=FALSE}
ggcorr(red[, 2:13], 
       palette = "RdGy",
       hjust = 0.9,
       size = 5,
       nbreaks = 6, 
       label_round = 2, 
       label = TRUE, 
       layout.exp = 3,
       label_size = 3) + 
  labs(title="Correlation of parameters")
```

First we check all the variables for correlation. We see which variables are
strongly corelated with each other. This function is found from this website
https://briatte.github.io/ggcorr/

Inspecting this, we can see alcohol level strongly correlated with quality, 
citric acid and sulphates also have slight positive correlation, volatile
acidity on the other hand has negative correlation. The other variables do not
have significant correlations. So if I want to make a model, I will most
probably use these 3 variables, alcohol level, volatile acidity and sulphates.



```{r}
ggplot(aes(factor(quality), 
            alcohol), 
        data = red) +
  geom_jitter( alpha = .3)  +
  geom_boxplot( alpha = .5,color = 'blue')+
  stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "red", 
               shape = 8, 
               size = 4) + 
   labs(title="Alcohol content ~ Quality of Wine")
```

I want to first investigate relation of alcohol content and quality of the wine.
I suspect there should be strong correlation.

Checking this graph we see good wines (quality>5) have more alcohol content.
Whereas wines that are not so good (quality<6) have varying alcohol content.

We can notice positive correlation between alcohol content and quality.


```{r}
ggplot(aes(factor(quality), 
            volatile.acidity), 
        data = red) +
  geom_jitter( alpha = .3)  +
  geom_boxplot( alpha = .5,color = 'blue')+
  stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "red", 
               shape = 8, 
               size = 4) + 
   labs(title="Volatile Acidity ~ Quality of Wine")
```

This shows the relationship between volatile acidity and quality. More volatile
 acidity causes the wine to taste like vinegar. We can see an opposite
 correlation between volatile acidity and quality.

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

As we have discussed earlier there is a strong corelation between perceived
quality of the wine and alcohol content and volatile acidity. Unexpected 
previously there is no corellation between the sugar content and free sulfur
level and quality of the wine, I was expecting a relationship between these
values.


### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

Here I noted a relationship between volatile acidity, fixed acidity and citric
acid level. Citric acid level is positively corelated with fixed acidity and 
negatively corelated with volatile acidity. So my observation is that wine
producers add citric acid to modify their acidity and lowering volatile acidity.
But some wine makers cannot do that (as discussed I found out that EU producers
cannot use citric acid, it would be good to check this.) Plotting some graphs
shows interesting results for wine that doesnot use any citric acid, those are
slightly worse in ratings as they contain more volatile acids.


### What was the strongest relationship you found?

Strongest relationships are the obvious ones such as pH and acid levels, total
sulfur vs free sulfur etc. But there are also some strong relationship that 
are not that obvious such as alcohol level and volatile acidity levels and 
quality of wine.


# Multivariate Plots Section


```{r echo=FALSE, Multivariate_Plot1}
ggplot(data = red, aes(volatile.acidity, alcohol, color = as.factor(quality))) +
  geom_point() +
  ylab('alcohol (% volume)') +
  scale_color_brewer(palette = 7) +
geom_smooth(method = "lm", se = FALSE,size=1)
```

I am first inspecting the effect of alcohol and Volatile acidity in quality of
wine. These two variables are the ones that have the most correlation value.
Alcohol positively correlates with quality of wine whereas, volatile acidity
negatively correlates. Plotting the graph we can immediately see the strong
relation of alcohol content and quality of the wine as the lines are stacked 
above each other according to the quality. On the other hand Volatile acidity
is more scattered. But it is still possible to see it is negatively correlated
with quality of wine, as the lines of better quality wine are located on the
left side of the graph. And finally by looking at the graph I though alcohol and
volatile acidity levels are positively correlated, but checking the correlations
table shows the opposite of this, these two variables are negatively correlated.
But at least we can see here for the best quality of wines, there is a strong
positive correlation.

```{r echo=FALSE, Multivariate_Plot2}
ggplot(data = red, aes(volatile.acidity, citric.acid, color=as.factor(quality))) +
  geom_point() +
  scale_color_brewer(palette = 7) +
  geom_smooth(method = "lm", se = FALSE,size=1)

```

Here this time I am looking at the relationship of volatile acidity and citric
acid levels with quality of wines. Here immediately we can see a very linear
strong negative correlation between citric acid and volatile acidity. And what
we can see here again is there are a lot of 0s for citric acid, as it is banned
in EU. It is not very clear but we can almost sense the negative correlation
of volatile acidity and wine quality as the bands of quality 6, 7, 8 are more to
the left of the 3, 4, 5.


```{r echo=FALSE, Multivariate_Plot3}
ggplot(data = red, aes(sulphates, alcohol, color=as.factor(quality))) +
  geom_point() +
  ylab('alcohol (% volume)') +
  xlab('Sulphates (% volume)') +
  scale_color_brewer(palette = 7) +
geom_smooth(method = "lm", se = FALSE,size=1)
```

The last observation is for alcohol and sulphates. Both have a positive 
correlation with the quality of a wine. We can see this as the color of dots
gets darker as we go towards to the up right corner compared to the bottom left
corner. 


# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

These plots strengtened mainly the two topics I have been discussing, quality
of wines and effect of alcohol and volatile acidity. And means of acidity (ie. 
volatile, fixed, citric) and quality. I conclude about the importance of the
volatile acidity on the quality. And lastly I have plotted the sulphates and 
alcohol in a graph to see their effect on the quality.

### Were there any interesting or surprising interactions between features?

The interesting conclusion from this study is that, wine is a very complex
beverage and there is no easy way to describe what makes a good wine. Playing 
with the parameters we noted certain correlations but apart from alcohol level
and volatile acidity, the affect of others are not so clear.

### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.


So as I discussed earlier I can use these 3 variables, alcohol level, volatile
acidity and sulphates. I think these 3 are enough to have a meaningful value to
make a linear model. I put the variables in linear model. The results show I
have an F value that is greater than 1 so there is a relationship. My r squared
value is 0.33 which is not very good but still it show a weak relationship.
I wont be spending too much time on this, as this is a topic for later course.
And wine quality is a very complicated task to fit into a model.

(referrence: https://www.r-bloggers.com/how-to-apply-linear-regression-in-r/)

```{r}
fit <- lm(red$quality ~ red$alcohol 
          + red$volatile.acidity + red$sulphates, data=red)
summary(fit) 

```
------

# Final Plots and Summary



### Plot One
```{r echo=FALSE, Plot_One}
ggplot(data=red, aes(x=quality)) +
  geom_histogram(binwidth = 1, color='black', fill='red') + 
  scale_x_continuous(breaks = seq(3, 8, 1)) +
  labs(title="Histogram of Wine Quality")+
  xlab("Wine Quality") +
  ylab("Number of Samples") +
  theme(text = element_text(size=12)) 

```

### Description One
This was the first plot I used in my study, as this is the main aspect that 
we are studying I put this in the concluding chapter. One can note from this 
graph majority of the wines are labeled as 5-6 rating whereas low quality or
high quality wines are lesser in numbers.

### Plot Two
```{r echo=FALSE, Plot_Two}

ggplot(data=red, aes(group=quality, x=quality, y=alcohol)) +
  geom_boxplot(color='black', fill='red') + 
  scale_x_continuous(breaks = seq(3, 8, 1))+
  xlab("Quality") +
  ylab("Alcohol (% volume)") +
  labs(title="Box Plot of Alcohol and Wine Quality")+
  theme(text = element_text(size=12)) 




```

### Description Two

Alcohol is strongly correlated with wine quality especially on the better wines
(Quality 7-8). This graph is showing this relationship. What I note is the
interesting case of wines with rating 5. They are the ones with the lowest mean.

### Plot Three
```{r echo=FALSE, Plot_Three}
ggcorr(red[, 2:13], 
       palette = "RdGy",
       hjust = 0.9,
       size = 5,
       nbreaks = 6, 
       label_round = 2, 
       label = TRUE, 
       layout.exp = 3,
       label_size = 3) + 
  labs(title="Correlation of parameters")

```

### Description Three

Here I put this plot as this summarises all the correlation of different 
parameters in one plot. There are some interesting things to note. At the start 
of the study I was certainly expecting some correlation between quality and 
residual sugar. But this plot and my previous examinations dont show any 
correlation. Also here it is easy to see the relationship between all the
acidity parameters.


------




# Reflection

This was a very interesting study and it gave me a good chance to practice 
all the functions I learned from R environment. Also I had a chance to learn
about red wines which will probably make me appreciate more of the taste of
red wine next time. There were some obvious outcomes and not so obvious ones 
for me as well. The level of alcohol and quality of the wine was surprising for
me. I was even expecting the opposite
as I think the more alcohol in the wine it would me more "bitter". 

The relationship of volatile acidity and quality of wine is less suprising. As 
generally people dont like the vinegar taste of the volatile acids.

This study can be extended with the addition of location of the wine, year it 
has passed and it is fermentation process as the other parameters would probably
be more useful. Such as less sulphates can make wines more susceptible to become
faulty.

Some hardships I had during this study is about the legends and axis, I
sometimes find it difficult to arrange the axis such that the information
is readable. But in general I find this study, quite successful, as I was
able to identify at least some parameters effecting wine taste, and maybe
next time I am enjoying my wine I will appreciate more.


### Resources

https://en.wikipedia.org/wiki/Acids_in_wine  
https://en.wikipedia.org/wiki/Wine_fault#Acetic_acid   
http://www.minitab.com/en-us/Published-Articles/Wine-Tasting-by-Numbers--Using-Binary-Logistic-Regression-to-Reveal-the-Preferences-of-Experts/   
https://briatte.github.io/ggcorr/   

